Bias correction in GLMs

### 2021-08-25 aprroximation_bias_correction.rmd
Autocommit

>>>>

```{r generate_pred2}
genpred2 <- function(mod, focal, level=0.95, steps=100, modelname="none", ...) {

	mm <- model.matrix(mod)
	quant <- seq(0, 1, length.out=steps)
	xvals <- as.vector(quantile(mm[, focal], quant))
	vnames <- vareffects:::get_vnames(mod)$termnames
	check <- !vnames %in% focal
	non.focal.terms <- vnames[check]
	betas <- coef(mod)
	non.focal.coefs <- betas[check]
	focal.coefs <- betas[!check]
	lp.non.focal <- as.vector(as.matrix(mm[, non.focal.terms, drop=FALSE]) %*% non.focal.coefs)
	lp.all <- unlist(lapply(xvals, function(x){
		mu <- mean(plogis(as.vector(c(x %*% focal.coefs) + lp.non.focal)))
		return(mu)
	}))

	out <- list(xx = xvals, xx2 = lp.all)
	return(out)
}

x1 <- genpred2(simple_mod, "wealthindex")
mean(x1$xx2)

### 2021-08-25 bias_correction.rmd
Autocommit

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

@@

Prediction of the variation in the response variable depends on whether the relationship between the response variable and the predictors is linear or nonlinear. For example, when response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$. This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative. Most packages for predicting responses make distribution assumptions about the predictors, (for example, conditioning at the mean value of the predictor), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging. We consider the following approaches for bias correction:

@@

	- Whole population
	- Quantiles

- Distributional conditioning
- Second-order correction

We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets.

@@

Coefficient plots

```{r simple_coef, fig.cap="Coefficient plot for simple logistic model."}
## True beta
true_beta_df <- data.frame(term=c("Intercept", "age", "wealthindex")
	, estimate=c(beta0, betaA, betaW)
)

## Tidy coef estimates
coef_df <- (broom::tidy(simple_mod, conf.int=TRUE)
#	%>% dotwhisker::by_2sd(sim_df)
	%>% mutate(term = gsub("\\(|\\)", "", term))
)
print(coef_df)

simple_coef_plot <- (plotEsize(coef_df) 
	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
	+ labs(colour="Type")
	+ scale_colour_manual(values=c("black", "blue"))
)
print(simple_coef_plot)
```

Effect sizes on logit scale:

```{r simple_logit}
coef_df_logit <- (coef_df
	%>% select(term, estimate, conf.low, conf.high)
	%>% group_by(term)
	%>% summarise_all(plogis)
)
print(coef_df_logit)
```

@@

We consider both \textbf{varpred} and population averaging approach; and then introduce bias correction to \textbf{varpred} predictions.

@@

- Age

```{r simple_vareffects_age} 
## varpred way
simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="Distribution")

@@

- Wealth index

```{r simple_vareff_wealthindex, fig.cap = "A comparison of various bias-correction approaches -- Distribution approach is based on averaging of non-focal predictors; Population approach involves using the observed values of non-focal predictors together with the quantiles of the focal predictors; while quantiles approach involves sampling (or picking quantiles of) the observed values of both focal and non-focal predictors. Except for the distribution approach, population and quantiles approaches give very close estimates."}
# Wealth index
## varpred 
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="Distribution")

@@

The observed population average is `r true_prop` while the estimated population averages (similar estimates with `varpred`) are:

@@

```{r reff_coef, fig.cap="Mixed model coefficient estimates."}
## Tidy coef estimates
reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
	%>% mutate(term = gsub("\\(|\\)", "", term))
	%>% filter(effect=="fixed")
)
print(reff_coef_df)

reff_coef_plot <- (plotEsize(reff_coef_df) 
	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
	+ labs(colour="Type")
	+ scale_colour_manual(values=c("black", "blue"))
)
print(reff_coef_plot)
```

Variable effect plots

- Age

```{r reff_vareffects_age}
## varpred way
reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="Distribution")

@@

- Wealth index

```{r reff_vareff_wealthindex, fig.cap = "Distribution based approach over-estimates the predictions in age and wealth index."} 
# Wealth index
## varpred
reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=FALSE, modelname="Distribution")
binned_df <- binfun(reff_mod, "wealthindex", "age")

@@

>>>>

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "", message=FALSE, warning=FALSE)

@@

Prediction of the variation in the response variable depends on whether the relationship between the response variable and the predictors is linear or nonlinear. For example, when response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$. This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative. Most packages for predicting responses make distribution assumptions about the non-focal predictors, (for example, conditioning at the mean value of the predictor), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging. We consider the following approaches for bias correction:

@@

	- Whole population
	- Quantiles
- Second-order correction

and provide a comparison with the uncorrected _Distributional conditioning_ (varpred or emmeans approach).

We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets -- for univariate and multivariate models.

We start with a univariate case where we have only one predictor.

@@

@@

@@

```{r simple_vareffects_age, echo=FALSE} 
## varpred way
simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="Distribution")

@@

```{r simple_vareff_wealthindex, echo=FALSE, fig.cap = "A comparison of variable predictions with uncorrected and corrected estimates. The uncorrected Distribution approach is based on averaging of non-focal predictors; Population approach involves using the observed values of non-focal predictors together with the quantiles of the focal predictors; while quantiles approach involves sampling (or picking quantiles of) the observed values of both focal and non-focal predictors. The uncorrected Distribution approach seems to over-predict and is slightly higher then the marginal estimates."}
# Wealth index
## varpred 
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="Distribution")

@@

The marginal mean is `r true_prop` while the estimated are:

@@

```{r reff_vareffects_age, echo=FALSE}
## varpred way
reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="Distribution")

@@

```{r reff_vareff_wealthindex, echo=FALSE, fig.cap = "Similar to the previous simple model, the quantile and population-based averaging gives very close estimates as opposed to the uncorrected distribution-based which appears to over-predict."} 
# Wealth index
## varpred
reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=FALSE, modelname="Distribution")
binned_df <- binfun(reff_mod, "wealthindex", "age")

@@

We can also add centered predictions as in the case of _varpred_:

```{r reff_vareffects_age_centered, echo=FALSE}
## varpred way
reff_vareff_age_cen <- varpred(reff_mod, "age", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
reff_vareff_age_quant_cen <- varpred(reff_mod, "age", isolate=TRUE, pop.ave="quantile", include.re=TRUE, modelname = "quantiles")
### Population
reff_vareff_age_pop_cen <- varpred(reff_mod, "age", isolate=TRUE, pop.ave="population", include.re=TRUE, modelname = "Population")
binned_df <- binfun(reff_mod, "age", "wealthindex", bins=15)

vareff_age <- reff_vareff_age_cen
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_quant_cen$preds, reff_vareff_age_pop_cen$preds))
age_plot_cen <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
```

```{r reff_vareff_wealthindex_cen, echo=FALSE, fig.cap = "Same plots above but with centered confidence bands..."} 
# Wealth index
## varpred
reff_vareff_wealthindex_cen <- varpred(reff_mod, "wealthindex", isolate=TRUE, modelname="Distribution")
binned_df <- binfun(reff_mod, "wealthindex", "age")

## Pop. average
### Quantiles
reff_vareff_wealthindex_quant_cen <- varpred(reff_mod, "wealthindex", isolate=TRUE, pop.ave="quantile", include.re=TRUE, modelname="quantiles")
### Population
reff_vareff_wealthindex_pop_cen <- varpred(reff_mod, "wealthindex", isolate=TRUE, pop.ave="population", include.re=TRUE, modelname="Population")

vareff_wealthindex <- reff_vareff_wealthindex_cen
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant_cen$preds, reff_vareff_wealthindex_pop_cen$preds))
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant_cen$preds, reff_vareff_wealthindex_pop_cen$preds))
wealthindex_plot_cen <- (plot(vareff_wealthindex)
 	+ labs(y="", colour="Model")
	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)

ggarrange(age_plot_cen, wealthindex_plot_cen, common.legend=TRUE)
```

\section{Multivariate model}

We now consider a model with multiple predictors.

```{r simple_sim_multi}
N <- 1e4
beta0 <- 0.7
betaA <- 0.2
betaE <- -0.7
betaW <- 0.5
age_max <- 1
age_min <- 0.2
expend_mean <- 0.2
expend_sd <- 1
age <- runif(N, age_min, age_max)
expenditure <- rnorm(N, expend_mean, expend_sd)
wealthindex <- rnorm(N, 0, 1)
eta <- beta0 + betaA * age + betaW * wealthindex + betaE*expenditure
sim_df <- (data.frame(age=age, expenditure=expenditure, wealthindex=wealthindex, eta=eta)
	%>% mutate(status = rbinom(N, 1, plogis(eta)))
	%>% select(-eta)
)
true_prop <- mean(sim_df$status)
print(true_prop)
```

```{r simple_mod_multi}
simple_mod_multi <- glm(status ~ age + expenditure + wealthindex, data = sim_df, family="binomial")
```

```{r simple_vareffects_age_multi, echo=FALSE}
## varpred way
simple_vareff_age_multi <- varpred(simple_mod_multi, "age", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
simple_vareff_age_quant_multi <- varpred(simple_mod_multi, "age", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
### Population
simple_vareff_age_pop_multi <- varpred(simple_mod_multi, "age", isolate=TRUE, pop.ave="population", modelname="Population")
binned_df <- binfun(simple_mod_multi, focal = "age", non.focal = c("expenditure", "wealthindex"), bins=50)

vareff_age <- simple_vareff_age_multi
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_quant_multi$preds, simple_vareff_age_pop_multi$preds))
age_plot_multi <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
```

```{r simple_vareffects_expenditure_multi, echo=FALSE}
## varpred way
simple_vareff_expenditure_multi <- varpred(simple_mod_multi, "expenditure", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
simple_vareff_expenditure_quant_multi <- varpred(simple_mod_multi, "expenditure", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
### Population
simple_vareff_expenditure_pop_multi <- varpred(simple_mod_multi, "expenditure", isolate=TRUE, pop.ave="population", modelname="Population")
binned_df <- binfun(simple_mod_multi, focal = "expenditure", non.focal = c("age", "wealthindex"), bins=50)

vareff_expenditure <- simple_vareff_expenditure_multi
vareff_expenditure$preds <- do.call("rbind", list(vareff_expenditure$preds, simple_vareff_expenditure_quant_multi$preds, simple_vareff_expenditure_pop_multi$preds))
expenditure_plot_multi <- (plot(vareff_expenditure)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=expenditure, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
```

```{r simple_vareffects_wealthindex_multi, echo=FALSE, fig.cap="A comparison of bias correction approaches for a simple multivariate logistic model."}
## varpred way
simple_vareff_wealthindex_multi <- varpred(simple_mod_multi, "wealthindex", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
simple_vareff_wealthindex_quant_multi <- varpred(simple_mod_multi, "wealthindex", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
### Population
simple_vareff_wealthindex_pop_multi <- varpred(simple_mod_multi, "wealthindex", isolate=TRUE, pop.ave="population", modelname="Population")
binned_df <- binfun(simple_mod_multi, focal = "wealthindex", non.focal = c("age", "expenditure"), bins=50)

vareff_wealthindex <- simple_vareff_wealthindex_multi
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_quant_multi$preds, simple_vareff_wealthindex_pop_multi$preds))
wealthindex_plot_multi <- (plot(vareff_wealthindex)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
ggarrange(age_plot_multi + rremove("ylab")
	, expenditure_plot_multi + rremove("ylab")
	, wealthindex_plot_multi + rremove("ylab")
	, nrow=1, common.legend=TRUE
)
```

\subsection{Mixed model}

```{r sim_mixed_multi}
beta0 <- 0.7
betaA <- 0.2
betaW <- 0.5
betaE <- -0.7
age_max <- 1
age_min <- 0.2
expend_mean <- 0.2
expend_sd <- 1

# Simulation parameters
nHH <- 100 # Number of HH (primary units) per year
nyrs <- 50 # Number of years to simulate
yrs <- 2000 + c(1:nyrs) # Years to simulate
N <- nyrs * nHH
## HH random effect sd
hhSD <- 0.5
# Generate dataset template
temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
		, years = rep(yrs, nHH)
		, age = runif(n=N, age_min, age_max)
		, expenditure = rnorm(N, expend_mean, expend_sd)
		, wealthindex = rnorm(n = N, 0, 1)
	)
)
# Simulate HH-level random effects (residual error)
hhRE <- rnorm(nHH, hhSD)
temp_df$hhRE <- hhRE[temp_df$hhid]
sim_df <- (temp_df
	%>% mutate(eta = beta0 + betaA * age + betaE * expenditure + betaW * wealthindex + hhRE
	, status = rbinom(N, 1, plogis(eta))
	)
	%>% select(-eta)
)
true_prop_reff <- mean(sim_df$status)
print(true_prop_reff)
```

```{r mixed_model_multi}
reff_mod_multi <- glmmTMB(status ~ age + expenditure + wealthindex + (1|hhid)
	, data = sim_df
	, family = binomial(link = "logit")
)
```

```{r reff_vareffects_age_multi, echo=FALSE}
## varpred way
reff_vareff_age_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
reff_vareff_age_quant_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
### Population
reff_vareff_age_pop_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, pop.ave="population", modelname="Population")
binned_df <- binfun(reff_mod_multi, focal = "age", non.focal = c("expenditure", "wealthindex"), bins=50)

vareff_age <- reff_vareff_age_multi
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_quant_multi$preds, reff_vareff_age_pop_multi$preds))
age_plot_multi <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
```

```{r reff_vareffects_expenditure_multi, echo=FALSE}
## varpred way
reff_vareff_expenditure_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
reff_vareff_expenditure_quant_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
### Population
reff_vareff_expenditure_pop_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="population", modelname="Population")
binned_df <- binfun(reff_mod_multi, focal = "expenditure", non.focal = c("age", "wealthindex"), bins=50)

vareff_expenditure <- reff_vareff_expenditure_multi
vareff_expenditure$preds <- do.call("rbind", list(vareff_expenditure$preds, reff_vareff_expenditure_quant_multi$preds, reff_vareff_expenditure_pop_multi$preds))
expenditure_plot_multi <- (plot(vareff_expenditure)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=expenditure, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
```

```{r reff_vareffects_wealthindex_multi, echo=FALSE, fig.cap="A comparison of bias correction approaches for a reff multivariate logistic model."}
## varpred way
reff_vareff_wealthindex_multi <- varpred(reff_mod_multi, "wealthindex", isolate=TRUE, modelname="Distribution")

## Pop. average
### Quantiles
reff_vareff_wealthindex_quant_multi <- varpred(reff_mod_multi, "wealthindex", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
### Population
reff_vareff_wealthindex_pop_multi <- varpred(reff_mod_multi, "wealthindex", isolate=TRUE, pop.ave="population", modelname="Population")
binned_df <- binfun(reff_mod_multi, focal = "wealthindex", non.focal = c("age", "expenditure"), bins=50)

vareff_wealthindex <- reff_vareff_wealthindex_multi
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant_multi$preds, reff_vareff_wealthindex_pop_multi$preds))
wealthindex_plot_multi <- (plot(vareff_wealthindex)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
	+ theme(legend.position="bottom")
)
ggarrange(age_plot_multi + rremove("ylab")
	, expenditure_plot_multi + rremove("ylab")
	, wealthindex_plot_multi + rremove("ylab")
	, nrow=1, common.legend=TRUE
)
```
